Supplementary Material (SM)

### 2020-04-18 .A White Noise Approach To Evolutionary Ecology/SM.tex
split main text and SM

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={A White Noise Approach to Evolutionary Ecology},
            pdfauthor={Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{hyperref}
\usepackage[left]{lineno}
\usepackage{amsmath}
\linenumbers
\usepackage{float}
\usepackage{flafter}

\title{A White Noise Approach to Evolutionary Ecology}
\author{Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon}
\date{3/1/2020}

\begin{document}
\maketitle
\begin{abstract}
We derive the dynamics of the distribution of a quantitative character
and the abundance of a biological population from a stochastic partial
differential equation driven by space-time white noise. In the process
we develop a useful set of heuristics to operationalize the powerful,
but abstract theory of white noise and measure-valued stochastic
processes. This approach allows us to compute the full implications of
demographic stochasticity on phenotypic distributions and abundances of
populations. We demonstrate the utility of our approach by deriving a
quantitative genetic model of diffuse coevolution mediated by
exploitative competition for a continuum of resources. Other than trait
and abundance distributions, this model predicts interaction networks
parameterized by rates of interactions, competition coefficients, and
selection gradients. We briefly investigate the relationship between
selection gradients and competition coefficients. This illustrative
investigation suggests absolute values of linear selection gradients are
negatively correlated with competition coefficients when the ratio of
interspecific trait variation to intraspecific trait variation is
approximately less than \(1.25\) and positively correlated otherwise.
The inverse relationship holds for quadratic selection gradients except
with threshold diversity ratio equal to unity instead of \(1.25\).
Hence, this approach contributes to the development of a synthetic
theory of evolutionary ecology by formalizing first principle
derivations of stochastic models that underlie rigorous investigations
of the relationship between feedbacks of biological processes and the
patterns of diversity they produce.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Our goal in this manuscript is to develop a rigorous, but accessible
approach to synthesize the stochastic dynamics of abundance, mean trait
and heritable variation in biological populations for the study of
theoretical evolutionary ecology. A primary aim of theoretical
evolutionary ecology is the development of mathematical approaches to
describe the evolution of populations and their interactions with both
the biotic and abiotic environments in which they are embedded. Given
this consideration, a natural scope for such an approach centers on
quantifying the abundance dynamics of populations and the evolution of
traits mediating their interactions as functions of relevant abiotic
factors. Although taking into account abundance, phenotype and
environment provides the basis for a partial understanding of the
complex nature of biological communities, a deeper understanding must
account for the effects of dispersal and the phylogeographic history of
interacting lineages (Kraft et al. 2007; Hickerson et al. 2010; Manceau,
Lambert, and Morlon 2016; McPeek 2017) along with the genetic basis of
ecologically relevant traits (Conner 2004; Fussman, Loreau, and Abrams
2007) and feedbacks between populations and the biogeochemical cycles
they ultimately depend on (Loreau 2010; Ågren and Andersson 2012). It is
therefore ideal that the development of any such mathematical approach
anticipates extensions to account for these important factors shaping
ecological communities, especially as empirical and conceptual work in
these directions continues to grow (Abdala-Roberts and Mooney 2014;
Kölzsch et al. 2015; Crutsinger 2015; Fitzpatrick et al. 2015, 2017;
Marx et al. 2017; Rudman et al. 2017; Skovmand et al. 2018; Nuland et
al. 2019; Harmon et al. 2019). Furthermore, the approach would benefit
from a stochastic component to capture the chance nature of biological
reality (Lande, Engen, and SÆther 2003; Meester et al. 2018; Mubayi et
al. 2019) and serve as a basis for the construction of statistical
methods that measure evolutionary and ecological processes occuring in
the wild. Such methods will tether theory to reality and allow for
rigorous tests of hypotheses on the structure and behavior of ecological
communities. In this paper we introduce a framework that establishes a
formal connection between the continuous-time dynamics of abundance and
quantitative traits in stochastically evolving populations. We then
demonstrate the utility of our framework through the derivation and
analysis of a model of diffuse coevolution and discuss how it can be
extended to account for the details mentioned above.

Current theoretical approaches to synthesize evolution and ecology have
capitalized on the fact that biological fitness plays a key role in
determining both sets of dynamics. While correlation of fitness and
genotype is the basis of evolution by natural selection, the mean
fitness across all individuals in a population determines the growth,
stasis or decline of abundance. In section \ref{deterministic} we review
the mathematical formalization of this connection, which has been
established in the contexts of population genetics (Crow and Kimura
1970; Roughgarden 1979), evolutionary game theory (Hofbauer and Sigmund
1998; Nowak 2006; Lion 2018), quantitative genetics (Lande 1982; Doebeli
1996; Lion 2018) and a unifying framework for these three distinct
approaches to evolutionary theory (Champagnat, Ferrière, and Méléard
2006) which includes our approach as a special case. We note here there
are two primary differences between the results presented in this paper
and those of Champagnat, Ferrière and Méléard (2006). First, instead of
developing a unifying approach to evolutionary theory, we focus on
developing a stochastic synthesis of population dynamics and
quantitative genetics including specific expressions for the dynamics of
abundance, mean trait and additive genetic variance. Second, although
Champagnat, Ferrière and Méléard (2006) provide rigorous proofs of
results relevant to our approach, they do not communicate the nessecary
tools we make use of for deriving the dynamics of abundance, mean trait
and additive genetic variance. We aim to communicate these tools to a
wide audience by striking a balance between intuitive appeal and
mathematical rigor.

Although reviewing previously established theoretical results reveals a
beautiful synthesis of evolution and population ecology, it also reveals
a gap in theoretical approaches to incorporate the intrinsically random
nature of populations. Specifically, in theoretical quantitative
genetics the derivation of a population's response to random genetic
drift is derived in discrete time under the assumption of constant
effective population size using arguments based on properties of random
samples (Lande 1976). Though this approach conveniently mimics the
formalism provided by the Wright-Fisher model of population genetics,
real population sizes fluctuate over time. Furthermore, since these
fluctuations are themselves stochastic, it seems natural to derive
expressions for the evolutionary response to demographic stochasticity
and consider how the results relate to characterizations of random
genetic drift. This can be done in continuous time for population
genetic models without too much technical overhead, assuming a finite
number of alleles (Lande, Engen, and Sæther 2009; Parsons, Quince, and
Plotkin 2010; Gomulkiewicz, Krone, and Remien 2017). However, for
populations with a continuum of types, such as a quantitative trait,
finding a formal approach to derive the evolutionary response to
demographic stochasticity has remained a vexing mathematical challenge.
In this paper we close this gap by combining the calculus of white noise
with results on rescaled limits of branching Brownian motion processes
(BBM) and stochastic partial differential equations (SPDE), which are
stochastic analogs of deterministic partial differential equations
(PDE). Our goal has two components: 1) Establish a novel synthetic
approach to theoretical evolutionary ecology that provides a formal
connection between demographic stochasticity and random genetic drift in
the context of quantitative traits. 2) Communicate some useful
properties of space-time white noise, BBM and SPDE to as wide of
audience as possible. With this goal in mind we will not provide a
rigorous treatment of any of these deep subjects. Instead, we introduce
a set of heuristics that only require the basic concepts of Reimann
integration, partial differentiation and some exposure to Brownian
motion and stochastic ordinary differential equations (SDE). For a
concise introduction to SDE and Brownian motion, we recommend the primer
by Evans (2014). Rigorous treatments of SPDE and rescaled limits of BBM
can be found in Walsh (1986) and Dawson (1993) respectively.

Through accomplishing these goals this work yields five main results.
Firstly, we develop a concise, but accurate approach for understanding
SPDE and provide heuristics to empower mathematical evolutionary
ecologists to derive novel models that relax key assumptions commonly
made in combining models of evolutionary and abundance dynamics. Second,
we formally derive general SDE driven by demographic stochasticity that
express the joint dynamics of abundance, mean trait and additive genetic
variance for populations exhibiting a wide range of potential phenotypic
distributions. Although these dynamics have been considered before while
accounting for demographic stochasticity (Polechová and Barton 2015), to
our knowledge this is the first time they have been rigorously derived
from first principles. Third, we simplify these SDE by approximating the
abundance density of phenotypes with a Gaussian curve and integrate the
results with a model of imperfect inheritance to find formula that
generalize classical quantitative genetic expressions. This includes an
expression for the response of additive genetic variance to selection in
terms of fitness gradients analogous to the expression for the response
in mean phenotype found by Lande (1976). Fourth, following classical
niche theory we provide a rigorous mechanistic derivation of fitness for
populations experiencing resource competition and abiotic stabilizing
selection using diffusion limits of measure-valued branching processes.
This calculation provides a demonstration for performing such diffusion
limits that can be applied in other contexts. Fifth, we explore
implications of the resulting model of diffuse coevolution by
investigating the relationship between selection gradients and
competition coefficients, discovering the conditions under which these
quantities positively or negatively covary.

We now summarize the basic structure of the paper. To provide motivation
for the stochastic equations developed later and background for our
model of coevolution, we begin with \S\ref{deterministic} by briefly
summarizing derivations of deterministic evolutionary and demographic
dynamics. Starting with a PDE (which we refer to as the Deterministic
Asexual Gaussian allelic model with Abundance dynamics, abbreviated
DAGA), we arrive at a general set of ordinary differential equations
modelling the dynamics of abundance, trait mean and trait variance. We
discuss the relationship between DAGA and previous models of phenotypic
evolution that have been employed in theoretical quantitative genetics.
In \S\ref{wnc} we discuss how a diffusion limit of a spatially
structured branching process (i.e., a measure-valued branching process)
leads to a SPDE that provides a natural stochastic generalization of
DAGA. We refer to this SPDE as the Stochastic Asexual Gaussian allelic
model with Abundance dynamics (abbreviated SAGA). The diffusion limit in
turn provides a rigorous method for constructing fitness functions used
in models of evolutionary ecology. However, before we discuss
measure-valued braching processes and their diffusion limits, we find it
necessary to review some fundamental results on the calculus of white
noise and introduce a set of mathematical tools based on these results.
In SM \S\ref{SDE_DERIV} we employ these tools to derive a system of SDE
generalizing our deterministic results to account for demographic
stochasticity. We find the effect of demographic stochasticity on the
evolution of mean trait and phenotypic variance characterizes the
process of random genetic drift in continuous time. However, although
biologically insightful, these equations remain difficult to analyze and
implement numerically. In \S\ref{equations} we use an assumption of
normally distributed trait values to simplify these expressions into
tractable formula that express the evolutionary response to selection in
terms of fitness gradients. We then account for the constraint of
adaptive evolution on the amount of heritable variation available in a
population by extending these results via a model of imperfect
inheritance. The resulting equations coincide with classical results in
quantitative genetics as a special case. In \S\ref{coev} we combine the
derived equations of population dynamics with classical niche theory to
formulate a model of coevolution across a guild of \(S\) species
participating in exploitative competition along a common resource
continuum. In SM \S\ref{diffuse} we apply a result on rescaled limits of
BBM to provide a rigorous derivation. In \S\ref{dynamics} we illustrate
our model of coevolution through numerical integration for \(S=100\)
species, tracking the dynamics of traits and abundances under scenarios
of weak and strong competition. We include an account of the natural
history of the simulated system and discuss the significance of
demographic stochasticity for structuring ecological communities. In
\S\ref{ecoevo} we provide expressions for selection gradients and
competition coefficients implied by our model and use these expressions
to investigate the relationship between the degree of competition and
strength of coevolution. Finally, \S\ref{conclusion} concludes with a
summary of accomplishments, a few remarks on the limits of this approach
and future directions to incorporate more explicitly the effects of
small populations, the genetic architecture of quantitative traits,
feedbacks with ecosystem processes and the macroevolutionary history of
interacting lineages.

\hypertarget{the-framework}{%
\section{The framework}\label{the-framework}}

At the core of our approach is a stochastic analog of the replicator
equation with mutation in continuous time and phenotypic space (Taylor
and Jonker 1978; Schuster and Sigmund 1983). From this stochastic
replicator-mutator equation we derive a system of SDE for the dynamics
of abundance, mean trait and additive genetic variance of a population.
Hence, our approach develops a quantitative genetic theory of
evolutionary ecology. A popular alternative to quantitative genetics is
the theory of adaptive dynamics. As demonstrated by Page and Nowak
(2002) and Champagnat, Ferrière and Méléard (2006), the canonical
equation of adaptive dynamics can be derived from the replicator-mutator
equation. Thus, one could start from the atomic roots of our approach
and pursue a stochastic adaptive dynamic theory instead. We choose the
former in anticipation of an extension of our approach that explicitly
models the genetic details of populations.

In this section we review the derivations of the replicator-mutator
equation and trait dynamics from abundance dynamics and extend these
formula along with related results to the stochastic case. The results
established in this section provide the framework from which larger
scale ecological stuctures, such as species abundance distributions and
interaction networks, can be computed.

\hypertarget{deterministic-dynamics}{%
\subsection{\texorpdfstring{Deterministic dynamics
\label{deterministic}}{Deterministic dynamics }}\label{deterministic-dynamics}}

Our review begins by considering the dynamics of an asexually
reproducing population in a homogeneous environment. For simplicity, we
first assume individuals are haploid and carry one of \(K\) alleles each
with a different fitness expressed as growth rate before introducing a
model involving a quantitative trait. Under these assumptions, the
derivation of the evolution of allele frequencies due to natural
selection can be derived from expressions of exponential growth. This,
and a few related approaches, have been provided by Crow and Kimura
(1970, \S 5.3). Specifically, denoting \(\nu_i\) the abundance of
individuals with allele \(i\) and \(m_i\) the growth rate of allele
\(i\) (called the Malthusian parameter in Crow and Kimura 1970), we have
\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation} Starting from this model, we get the total abundance of
the population as \(N=\sum_{i=1}^K\nu_i\), the frequency of allele \(i\)
as \(p_i=\nu_i/N\) and the mean fitness of the population as
\(\bar m=\sum_{i=1}^Kp_im_i\). Hence, we can employ some elementary
calculus to derive the dynamics of abundance \(dN/dt\) and the dynamics
of allele frequencies \(dp_1/dt,\dots,dp_K/dt\) as \begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation} \begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation} Two important observations of these equations include (i)
mean fitness \(\bar m\) determines the abundance dynamics of the entire
population and (ii) allele \(i\) will increase (decrease) in frequency
if \(m_i>\bar m\) (\(<\bar m\)). Equation (\ref{popgen}) is known in the
field of evolutionary game theory as the replicator equation (Hofbauer
and Sigmund 1998; Nowak 2006; Lion 2018; Taylor and Jonker 1978;
Schuster and Sigmund 1983). Instead of being explicitly focused on
alleles, the replicator equation describes the fluctuations of relative
abundances of various \emph{types} in a population in terms of the vital
rates of each type. Using a matrix of transition rates between differing
types, it is straight-forward to extend the replicator equation to
include mutation, which is known as the replicator-mutator equation
(Nowak 2006).

Inspired by equations (\ref{first})-(\ref{popgen}), we derive an analog
of the replicator-mutator equation for a continuum of types (that is,
for a quantitative trait). In particular, we model a continuously
reproducing population with trait values \(x\in\mathbb{R}\) and an
abundance density \(\nu(x,t)\) that represents the amount of individuals
in the population with trait value \(x\) at time \(t\). Hence, the
abundance density satisfies \(N(t)=\int_{-\infty}^{+\infty}\nu(x,t)dx\)
and \(p(x,t)=\nu(x,t)/N(t)\) is the relative density of trait \(x\)
which we also refer to as the phenotypic distribution.

To stay within the realm of biological plausibility we require a set
technical assumptions. First, we assume the initial abundance density is
continuous, non-negative, integrable and has finite trait mean and
variance. That is, we assume \(\nu(x,0)\) is continuous in \(x\),
satisfies \(\nu(x,0)\geq0\) for all \(x\in\mathbb{R}\) and
\begin{equation}
N(0)=\int_{-\infty}^{+\infty}\nu(x,0)dx<+\infty,
\end{equation} \begin{equation}
-\infty<\bar x(0)=\int_{-\infty}^{+\infty}xp(x,0)dx<+\infty,
\end{equation} \begin{equation}
\sigma^2(0)=\int_{-\infty}^{+\infty}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation} where \(\bar x(t)\) and \(\sigma^2(t)\) are respectively
the mean trait and phenotypic variance at time \(t\geq0\). Second, we
assume selection is determined by the growth rate \(m(h,x)\) that is
differentiable with respect to both arguments and is bounded above by
some \(r\in\mathbb{R}\) for all \(x\in\mathbb{R}\) and \(h\geq0\). We
also assume \((1+h)C\leq m(h,x)\) for some constant \(C>0\) so that
\(m(h,x)\) is linear or more than linear in \(h\). We denote by \(K\) an
operator that takes functions such as \(\nu(x,t)\) as its argument and
returns other functions that may still depend on the spatial variable
\(x\) and the temporal variable \(t\). For our application we will be
focused on operators \(K\) that can be written as
\((K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x-y)\nu(y,t)dy\) for some
non-negative and bounded function \(\kappa\). We set \(h=(K\nu)(x,t)\)
to account for nonlocal effects of the abundance density \(\nu(x,t)\) on
the fitness of individuals with trait value \(x\) (Volpert 2014). As an
example of a nonlocal effect, consider a population experiencing
intraspecific competition for a common resource such that the
competitive interactions are not mediated by any set of traits. In this
case all individuals in the population effect the fitness of one another
even though they may exhibit a diverse range of trait values. The term
nonlocal comes from interpreting the trait value as spatial location.
From here on we abbreviate \(m(h,x)=m\big((K\nu)(x,t),x\big)\) to just
\(m(\nu,x)\). Third, we assume mutation is captured by diffusion with
coefficient \(\frac{\mu}{2}\). With these technicalities addressed we
model the demographic dynamics of a population and the dynamics of a
quantitative character simultaneously by the PDE
\begin{equation}\label{eq1}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)
\end{equation} with the initial condition \(\nu(x,0)\) described above.
Equation (\ref{eq1}) is a semilinear parabolic PDE and is referred to as
a scalar reaction-diffusion equation (Zheng 2004; Evans 2010). To ensure
existence and uniqueness of solutions to (\ref{eq1}), we further assume
that for each \(M>0\) there exists a constant \(L_M>0\) depending on
\(M\) such that
\(\int_\mathbb{R}\nu_1(x)dx,\int_\mathbb{R}\nu_2(x)dx\leq M\) implies
\begin{equation}\label{local_lipschitz}
\int_\mathbb{R}\left|m(\nu_1,x)\nu_1(x)-m(\nu_2,x)\nu_2(x)\right|dx\leq L_M\int_\mathbb{R}|\nu_1(x)-\nu_2(x)|dx.
\end{equation} In SM \S\ref{finite} we combine the above assumptions on
growth rate, mutation and initial condition to prove that solutions to
(\ref{eq1}) satisfy \(N(t),|\bar x(t)|,\sigma^2(t)<+\infty\) for all
\(t\geq0\).

Interpreting the trait value \(x\) as spatial location and ignoring
nonlocal effects, which can be satisfied by setting \(K\nu=\nu\), this
model of spatially distributed population dynamics has been intensely
studied (for a review, see Cantrell and Cosner 2004). When \(\mu=0\),
equation (\ref{eq1}) can be seen as an analog of equation (\ref{first})
for a continuum of types. By assuming mutation acts as diffusion, the
effect of mutation causes \(\nu(x,t)\) to flatten out over time. In
fact, if the growth rate is constant across \(x\), then this model of
mutation will cause \(\nu(x,t)\) to converge to a flat line as
\(t\to\infty\).

Although clearly an idealized representation of biological reality, this
model is sufficiently general to capture a large class of dynamics
including density dependent growth and frequency dependent selection. As
an example, logistic growth combined with quadratic stabilizing
selection can be captured using the growth rate
\begin{equation}\label{special_m}
m(\nu,x)=r-\frac{a}{2}(\theta-x)^2-c\int_{-\infty}^{+\infty}\nu(y,t)dy
\end{equation} where \(r\in\mathbb{R}\) is the intrinsic rate of growth
in the absence of abiotic stabilizing selection, \(a>0\) is the strength
of abiotic stabilizing selection around the phenotypic optimum
\(\theta\in\mathbb{R}\), \(c>0\) captures the sensitivity of fitness to
intraspecific competition and \(\kappa(x-y)=1\) so that competitive
interactions cause the same reduction in fitness regardless of trait
value. This exemplary fitness function has a few convenient properties.
First, the effect of competition induces a local carrying capacity on
the population, leading to a finite equilibrium abundance over bounded
subsets of \(\mathbb{R}\). Second, abiotic selection prevents the
abundance density from diffusing too far from the abiotic optimum,
leading to a finite equilibrial abundance across all of \(\mathbb{R}\),
\(\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})\),
finite equilibrial mean trait \(\lim_{t\to\infty}\bar x(t)=\theta\) and
finite equilibrial phenotypic variance
\(\lim_{t\to\infty}\sigma^2(t)=\sqrt{\frac{\mu}{a}}\), so long as
\(\bar x(0)\in\mathbb{R}\), \(\sigma^2(0)\in[0,+\infty)\),
\(N(0)\in(0,+\infty)\) and \(r>\frac{1}{2}\sqrt{a\mu}>0\). We prove
these statements in SM \S\ref{equilib}. Note in particular the
coincidence of equilibrial phenotypic variance predicted by this model
and that predicted by the classic Gaussian allelic model of quantitative
genetics (Lande 1975; Bürger 2000; Johnson and Barton 2005). In
\S\ref{coev} and SM \S\ref{diffuse} we find this fitness function is
closely related to a fitness function derived from niche theory that we
use to model coevolution driven by resource competition.

To derive a replicator-mutator equation from equation (\ref{eq1}), we
employ the chain rule from calculus. Writing
\(\bar m(t)=\int_{-\infty}^{+\infty}m(\nu,x)p(x,t)dx\) for the mean
fitness, we have

\begin{multline}\label{N_det}
\frac{d}{dt}N(t)=\frac{d}{dt}\int_{-\infty}^{+\infty}\nu(x,t)dx=\int_{-\infty}^{+\infty}\frac{\partial}{\partial t}\nu(x,t)dx \\
=\int_{-\infty}^{+\infty} m(\nu,x)\nu(x,t)dx+\int_{-\infty}^{+\infty}\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)dx \\
=N(t)\int_{-\infty}^{+\infty} m(\nu,x)p(x,t)dx=\bar m(t) N(t).
\end{multline} Using our assumptions on mutation and rate of growth, we
show in SM \S\ref{finite} \(\nu(x,t)\) is twice differentiable with
respect to \(x\) and \(\int_{-\infty}^{+\infty}\nu(x,t)dx<\infty\) for
all \(t\geq0\). This implies that we are justified in swapping the order
of differentiation and integration and the result
\(\int_{-\infty}^{+\infty}\frac{\partial^2}{\partial x^2}\nu(x,t)dx=0\)
can be derived from the fundamental theorem of calculus. Biological
reasoning agrees with this latter result since mutation neither creates
nor destroys individuals, but merely changes their type from their
parental type. Taking the same approach, we derive the dynamics of the
phenotypic distribution \(p(x,t)\) in response to selection and muation
as

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=\Big(m(\nu,x)-\bar m(t)\Big) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline} This result closely resembles Kimura's
continuum-of-alleles model (Kimura 1965; Bürger 1986). The primary
difference being that our model utilizes diffusion instead of
convolution with an arbitrary mutation kernel. Of course, our model of
mutation can be derived as an approximation to Kimura's model, which,
for sexually reproducing populations, has been referred to as the
Gaussian allelic approximation in reference to the distribution of
mutational effects at loci in a genome on the values of traits (Lande
1975; but see also Bürger 2000; along with Johnson and Barton 2005), the
infinitesimal model in reference to modelling continuous traits as being
encoded by an infinite number of loci each having infinitesimal effect
(Barton, Etheridge, and Véber 2017) and the Gaussian descendants
approximation in reference to offspring trait values being normally
distributed around their parental values (Turelli 2017). Alternatively,
since diffusion is the continuous-time equivalent to convolution against
a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can
also be seen as a special case of Kimura's continuum-of-alleles model.
We return to these concepts in \S\ref{inheritance} where we review a
simple model of inheritance that leads us to an expression for the
stochastic evolution of additive genetic variance. To distinguish this
model from previous models of phenotypic evolution we refer to PDE
(\ref{eq1}) from which (\ref{p_PDE}) was derived as the Deterministic
Asexual Gaussian allelic model with Abundance dynamics (abbreviated
DAGA). Later, we will extend this model to include the effects of
demographic stochasticity, which we refer to as the Stochastic Asexual
Gaussian allelic model with Abundance dynamics (abbreviated SAGA).

The covariance of fitness and phenotype across the population is defined
as \begin{equation}
\mathrm{Cov}_t\Big(m(\nu,x),x\Big)=\int_{-\infty}^{+\infty}\Big(m(\nu,x)-\bar m(t)\Big)\Big(x-\bar x(t)\Big)p(x,t)dx.
\end{equation} Hence, the dynamics of the mean trait \(\bar x(t)\) can
be derived as

\begin{multline}\label{detmean}
\frac{d}{dt}\bar x(t)=\frac{d}{dt}\int_{-\infty}^{+\infty} x p(x,t)dx=\int_{-\infty}^{+\infty} x\frac{\partial}{\partial t} p(x,t)dx \\
=\int_{-\infty}^{+\infty} x\big(m(\nu,x)-\bar m(t)\big) p(x,t)+x\frac{\mu}{2}\frac{\partial^2}{\partial x^2} p(x,t)dx
=\mathrm{Cov}_t\Big(m(\nu,x),x\Big).
\end{multline}

Equation (\ref{detmean}) is a continuous time analog of the well known
Robertson-Price equation without transmission bias (Robertson 1966;
Price 1970; Frank 2012; Queller 2017; Lion 2018). Whether or not the
covariance of fitness and phenotype creates change in \(\bar x\) to
maximize mean fitness \(\bar m\) depends on the degree to which
selection is frequency dependent (Lande 1976). Since this change is
driven by a covariance with respect to phenotypic diversity, the
response in mean trait to selection is mediated by the phenotypic
variance. In particular, when \(\sigma^2=0\), \(\bar x\) will not
respond to selection. The result
\(\int_{-\infty}^{+\infty} x\frac{\partial^2}{\partial x^2} p(x,t)dx=0\)
can be found by applying integration by parts. Following the approach
taken to calculate the evolution of \(\bar x\), we find the response of
phenotypic variation to this model of selection and mutation is
\begin{equation}\label{vardyn_det}
\frac{d}{dt}\sigma^2(t)=\mathrm{Cov}_t\Big(m(\nu,x),(x-\bar x)^2\Big)+\mu.
\end{equation} For the sake of space we relegate the derivation of
\(d\sigma^2/dt\) to SM \S\ref{var_deriv}. In the absence of mutation
equation (\ref{vardyn_det}) agrees with the result derived by Lion
(2018) for discrete phenotypes. From a statistical perspective, if we
think of \((x-\bar x)^2\) as a square error, then in analogy to the
dynamics of the mean trait, we see that the response in \(\sigma^2\) to
selection can be expressed as a covariance of fitness and square error,
which is defined in analogy to \(\mathrm{Cov}_t(m(\nu,x),x)\). Just as
for the evolution of \(\bar x(t)\), this covariance also creates change
in \(\sigma^2\) that can either increase or decrease mean fitness
\(\bar m\), depending on whether or not individual fitness depends on
the distribution of traits across the entire population. The effect of
selection on phenotypic variance can be positive or negative depending
on whether selection is stabilizing or disruptive.

>>>>

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Supplementary Material (SM)},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{xr}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{hyperref}
\usepackage[left]{lineno}
\usepackage{amsmath}
\linenumbers
\externaldocument{draft}
\usepackage{float}
\usepackage{flafter}

\title{Supplementary Material (SM)}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
